Hypercoercions and a Framework for \\ Equivalence of Cast Calculi

### 2019-12-29 paper.tex
replace observation with observable to follow the convention in Jeremy 2012

\newcommand{\mbind}{\ensuremath{\underline{\gg}}}

@@

Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
machine and Fig.~\ref{fig:state} gives a grammar for machine states
$s$, including a definition of values $v$ and value typing. 
%Fig.~\ref{fig:state} gives a grammar for machine states
%$s$, including a definition of values and value typing and
%Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
%machine.
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions, (1) those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation, and (2) those that return a value to the
current continuation and possibly perform a computation.
Corresponding to (1) and (2), the machine state is either in an evaluating
$\sOOinspect{e}{\mathcal{E}}{\kappa}$ or continuing $\sOOreturn{v}{\kappa}$ configuration, 
respectively. Additionally, there is the $\sOOhalt{o}$ configuration, where the 
machine halts with an observation ($o$). Observations include all value 
constructors and blame. The function converting values to 
observations ($observe(v) = o$) is defined in the obvious way.
%Let $s$ range over machine states. A state is either looking at an 
%expression to decide what to do next, returning a value to a continuation, or 
%halting with an observation.

@@

\begin{figure}
  Machine state and other runtime data structures
  \[
  \begin{array}{lrcl}
  \stxrule{\mathcal{E}}{Environments}{
  	\text{a partial function}\;\{\langle x , v \rangle , \dots \}
%  	\text{partial functions that map  variables to values}
  }
  \stxrule{v}{Values}{
    \vOOtt{} \mid
    \vOOfun{\mathcal{E}}{x}{e} \mid
    \vOOcons{v}{v} \mid
    \vOOinl{v} \mid
    \vOOinr{v} \mid   
    \vOOcast{v}{c}
  }
  \stxrule{c}{Casts}{
    \cOOcast{T}{l}{T}
  }
  \stxrule{I}{Injectable types (\lazyD)}{
    P
  }
  \stxrule{I}{Injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\star}{\star} \mid
    \star \times \star \mid
    \star + \star
  }
  \stxrule{o}{Observations}{
    \oOOinj \mid
    \oOOsole \mid
    \oOOfun \mid
    \oOOcons \mid
    \oOOinl \mid
    \oOOinr \mid
    \oOOblame{l}
  }
  \stxrule{r}{Cast results}{
    \rOOsucc{v} \mid
    \rOOfail{l}
  }
  \stxrule{s}{States}{
    \sOOinspect{e}{\mathcal{E}}{\kappa} \mid{}
    \sOOreturn{v}{\kappa} \mid{}
    \sOOhalt{o}
  }
  \stxrule{\kappa}{Continuations}{
    \kOOmt \mid
    \kOOconsI{T}{T}{e}{\mathcal{E}}{\kappa} \mid
    \kOOconsII{T}{T}{v}{\kappa} \mid
    \kOOinl{T}{T}{\kappa}
  }
  \stxrulecont{
    \kOOinr{T}{T}{\kappa} \mid
    \kOOappI{e}{\mathcal{E}}{\kappa} \mid
    \kOOappII{v}{\kappa} \mid
    \kOOcar{\kappa}
  }
  \stxrulecont{ 
    \kOOcdr{\kappa}\mid
    \kOOcaseI{e}{e}{\mathcal{E}}{\kappa}\mid
    \kOOcaseII{v}{e}{\mathcal{E}}{\kappa}
  }
  \stxrulecont{
    \kOOcaseIII{v}{v}{\kappa} \mid
    \kOOcast{c}{\kappa}
  }
  \end{array}
  \]

@@

%% Three observations on this definition leads to hypercoercion: 
%% \begin{enumerate}
%%  \item The length of normal coercion is at most three.
%%  \item Projections are always at the beginning when present.
%%  \item Injections and failures are always at the end when present.
%% \end{enumerate}

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercoercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the no-op,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the no-op.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

@@

\begin{proposition}[Equivalence of Two \lazyD{} Cast ADTs]
  \label{thm:surely-lazyD-eqv}
  Assume $C_1$ and $C_2$ implements Lazy D Cast ADT,
  \[
  eval_{\mathcal{S}(C_1)} = eval_{\mathcal{S}(C_2)}.
  \]
\end{proposition}
\begin{proof}
  Without loss of generality, assume $\judgeSreduce{C_1}{e}{o}$ and prove 
  $\judgeSreduce{C_2}{e}{o}$.\\
  By the definition of $eval_{\mathcal{S}(C)}$, we know 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  By induction on this reduction sequence and Lemma~\ref{thm:CEKS-bisim}, we 
  have 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  that is, $\judgeSreduce{C_2}{e}{o}$
\end{proof}

@@

\begin{theorem}[\effCEK{C} is Correct wrt. \ineffCEKD{}]
  \label{thm:surely-lazyD-correct}
  Suppose $C$ is a \lazyD{} Cast ADT.  If
  $\judgetype{\emptyset}{e}{T}$ and $o : T$, then
  \[
  \evalEqv{\effCEK{C}}{\ineffCEKD}
%  \judgeSeval{C}{e}{o}
%  \;\text{if and only if}\;
%  \judgeDeval{e}{o}
  \]
\end{theorem}
\begin{proof}
	Immediately from 
	Equivalence of \lazyD{} Cast ADTs (Lemma~\ref{thm:surely-lazyD-eqv})
%	\evalEqv{\effCEK{C}}{\effCEK{L}} (Lemma~\ref{thm:surely-lazyD-eqv})
	and
	$L$ is Correct wrt. \ineffCEKD{} (Lemma~\ref{lem:L-correct})
%	\evalEqv{\effCEK{L}}{\ineffCEKD} (Lemma~\ref{lem:L-correct})
\end{proof}

>>>>

\newcommand{\mbind}{\ensuremath{\;\underline{\gg}\;}}

@@

Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
machine and Fig.~\ref{fig:state} gives a grammar for machine states
$s$, including a definition of values $v$ and value typing. 
%Fig.~\ref{fig:state} gives a grammar for machine states
%$s$, including a definition of values and value typing and
%Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
%machine.
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions, (1) those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation, and (2) those that return a value to the
current continuation and possibly perform a computation.
Corresponding to (1) and (2), the machine state is either in an evaluating
$\sOOinspect{e}{\mathcal{E}}{\kappa}$ or continuing $\sOOreturn{v}{\kappa}$ configuration, 
respectively. Additionally, there is the $\sOOhalt{o}$ configuration, where the 
machine halts with an observable ($o$). Observables include all value 
constructors and blame. The function converting values to 
observables ($observe(v) = o$) is defined in the obvious way.

@@

\begin{figure}
  Machine state and other runtime data structures
  \[
  \begin{array}{lrcl}
  \stxrule{\mathcal{E}}{Environments}{
  	\text{a partial function}\;\{\langle x , v \rangle , \dots \}
%  	\text{partial functions that map  variables to values}
  }
  \stxrule{v}{Values}{
    \vOOtt{} \mid
    \vOOfun{\mathcal{E}}{x}{e} \mid
    \vOOcons{v}{v} \mid
    \vOOinl{v} \mid
    \vOOinr{v} \mid   
    \vOOcast{v}{c}
  }
  \stxrule{c}{Casts}{
    \cOOcast{T}{l}{T}
  }
  \stxrule{I}{Injectable types (\lazyD)}{
    P
  }
  \stxrule{I}{Injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\star}{\star} \mid
    \star \times \star \mid
    \star + \star
  }
  \stxrule{o}{Observables}{
    \oOOinj \mid
    \oOOsole \mid
    \oOOfun \mid
    \oOOcons \mid
    \oOOinl \mid
    \oOOinr \mid
    \oOOblame{l}
  }
  \stxrule{r}{Cast results}{
    \rOOsucc{v} \mid
    \rOOfail{l}
  }
  \stxrule{s}{States}{
    \sOOinspect{e}{\mathcal{E}}{\kappa} \mid{}
    \sOOreturn{v}{\kappa} \mid{}
    \sOOhalt{o}
  }
  \stxrule{\kappa}{Continuations}{
    \kOOmt \mid
    \kOOconsI{T}{T}{e}{\mathcal{E}}{\kappa} \mid
    \kOOconsII{T}{T}{v}{\kappa} \mid
    \kOOinl{T}{T}{\kappa}
  }
  \stxrulecont{
    \kOOinr{T}{T}{\kappa} \mid
    \kOOappI{e}{\mathcal{E}}{\kappa} \mid
    \kOOappII{v}{\kappa} \mid
    \kOOcar{\kappa}
  }
  \stxrulecont{ 
    \kOOcdr{\kappa}\mid
    \kOOcaseI{e}{e}{\mathcal{E}}{\kappa}\mid
    \kOOcaseII{v}{e}{\mathcal{E}}{\kappa}
  }
  \stxrulecont{
    \kOOcaseIII{v}{v}{\kappa} \mid
    \kOOcast{c}{\kappa}
  }
  \end{array}
  \]

@@

%% Three observables on this definition leads to hypercoercion: 
%% \begin{enumerate}
%%  \item The length of normal coercion is at most three.
%%  \item Projections are always at the beginning when present.
%%  \item Injections and failures are always at the end when present.
%% \end{enumerate}

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observable that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercoercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the no-op,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the no-op.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

@@

\begin{proposition}[Equivalence of Two \lazyD{} Cast ADTs]
  \label{thm:surely-lazyD-eqv}
  Assume $C_1$ and $C_2$ implements Lazy D Cast ADT. If 
  $\judgetype{\emptyset}{e}{T}$ and $o : T$ then
  \[
  \evalEqv{\effCEK{C_1}}{\effCEK{C_2}}
  \]
%  \[
%  eval_{\mathcal{S}(C_1)} = eval_{\mathcal{S}(C_2)}.
%  \]
\end{proposition}
\begin{proof}
  Without loss of generality, assume $\judgeSeval{C_1}{e}{o}$ and prove 
  $\judgeSeval{C_2}{e}{o}$.\\
  By the definition of $eval_{\mathcal{S}(C)}$, we know 
\[
\judgeSreduceTrans{C_1}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  By induction on this reduction sequence and Lemma~\ref{thm:CEKS-bisim}, we 
  have 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  that is, $\judgeSeval{C_2}{e}{o}$
\end{proof}

@@

\begin{theorem}[\effCEK{C} is Correct wrt. \ineffCEKD{}]
  \label{thm:surely-lazyD-correct}
  Suppose $C$ is a \lazyD{} Cast ADT.  If
  $\judgetype{\emptyset}{e}{T}$ and $o : T$ then
  \[
  \evalEqv{\effCEK{C}}{\ineffCEKD}
  \]
\end{theorem}
\begin{proof}
	Immediately from 
	Equivalence of \lazyD{} Cast ADTs (Lemma~\ref{thm:surely-lazyD-eqv})
%	\evalEqv{\effCEK{C}}{\effCEK{L}} (Lemma~\ref{thm:surely-lazyD-eqv})
	and
	$L$ is Correct wrt. \ineffCEKD{} (Lemma~\ref{lem:L-correct})
%	\evalEqv{\effCEK{L}}{\ineffCEKD} (Lemma~\ref{lem:L-correct})
\end{proof}
